A regional scale assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2019-10-22 manuscript/chapter_1.tex
Added some model outputs

A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}. The strength and direction of these observed relationships varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{}. Over the past decade, many observational studies of the Biodiversity-Ecosystem Function Relationship (BEFR) have been conducted in tropical and temperate forests, and grasslands \citep{Chen2011}. The representation of dry tropical ecosystems in the BEFR literature however, is poor. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in Costa Rica.

In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity. Resource extraction directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation dynamics and affecting climate seasonality \citep{}. Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning \textgreater{}4 million km\textsuperscript{2}. The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}. They have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the banner of REDD+ \citep{Hinsley2015}. A small number of academic studies in these woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2017, and, others}, but all have been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends cannot be reliably inferred. \citet{Fayolle2018}

The role of environmental factors in shaping the BEFR generally is poorly understood \citep{Ratcliffe2017}. Climatic variation is known to affect woody biomass \citep{Michaletz2014}, so it is important to account for these factors and acknowledge their interaction with woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, which is presumably positively correlated with biomass, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum. However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.

\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under certain environmental conditions. In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}. Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}. Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic rates. Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}. In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas. The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it. Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition. Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness \textit{per se}.

@@

All stems \textgreater{}5 cm DBH were measured within each plot. For each tree, species, DBH and height to the highest branch material were recorded. Height was measured through a variety of means \todo{audit of height methods}. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial, parameters estimated using a test dataset \citep{Ryanunbublished}.

@@

Plots were assigned to biogeographical groups after \citet{Fayolle2018}. Groups were identified using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination. Plot data used in this study occurred in five biogeographical groups, here referred to as: \todo{AS WHAT?}.

@@

Structural Equation Models (SEM) investigated the regional determinants of woody Above-Ground Biomass (AGB). All SEMs were constructed and analysed in the \textit{lavaan} package in R version 3.6.0. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}. A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is known to affect productivity, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, and soil composition. Indepedently of total precipitation, precipitation seasonality (measured as the covariance of daily precipitation) determines, simplistically, whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth. We specified composite variables for moisture availability, temperature suitability, soil fertility, fire regime, tree species diversity and tree structural diversity. Latent variables. Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to account for measurement error.

A conceptual model was defined based on \textit{a priori} hypotheses of the direct and indirect effects of abiotic and biotic factors on AGB (Figure \ref{con_mod}). Latent variables were defined based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity, and tree structural diversity. Observed variables were collected in order to best inform these latent constructs.

Observed variables of the latent variables were standardized to Z-scores for analysis. Standardization puts latent variables on the same scale, with a mean of zero and a standard deviation of one. It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates problems arising from the observed variables being on different scales. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix.

The factor loadings of each observed variable in a latent variable were set to 1 in all cases, assuming that each has an equal contribution to its latent variable. Due to some latent variables being regressed against both structural diversity and AGB, factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM.

@@

The model allowed us to compare the relative effect sizes of the latent variables in determining AGB, and to assess the indirect effects of abiotic environment on AGB via species and structural tree diversity. Models including either abiotic or biotic factors only were compared using Comparative Fit Index (CFI), Akaike Information Criterion (AIC) and the R\textsuperscript{2} coefficient of determination for AGB.

Two sets of SEMs were fitted. The first used only tree diversity latent constructs to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship. The latent variable loadings differed between biogeographic clusters of plots. The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and biomass.

@@

Simple correlations showed a positive association between Aboveground Biomass (AGB) and a composite latent variable representing tree species diversity ($r_{(1031)}$ = 0.30, $t$ = 11.3).

In a simple path model describing the indirect effect of tree species diversity on AGB via the mediating effect of stand structural diversity, species diversity had a negative effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}). Stand structural diversity had a strong positive effect on AGB (\todo{Test stats}).

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes}
	\caption{Path coefficients (slopes) for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Path coefficients are +/- 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be non-significant effects.}
	\label{struc_model_slopes}
\end{figure}

@@

The latent variable of tree species diversity had a negative direct effect on AGB across the region.

@@

>>>>

A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}. The strength and direction of these observed relationships varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the Biodiversity-Ecosystem Function Relationship (BEFR) have been conducted in tropical and temperate forests, and grasslands \citep{Chen2011}. The representation of dry tropical ecosystems in the BEFR literature however, is poor. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the highly endemic wet tropical forests Costa Rica.

In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity. Resource extraction directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}. Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning \textgreater{}4 million km\textsuperscript{2}. The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}. These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the banner of REDD+ \citep{Hinsley2015}. A small number of academic studies in these woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2017, and, others}, but all have been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends cannot be reliably inferred. \citet{Fayolle2018}

The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}. Climatic variation is known to affect woody biomass \citep{Michaletz2014}, so it is important to account for these factors and acknowledge their interaction with both woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum. However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.

\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions. In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}. Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.

Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic rates. Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}. In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas. The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it. Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition. Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness \textit{per se}.

@@

All stems \textgreater{}5 cm DBH were measured within each plot. For each tree, species, DBH and height to the highest branch material were recorded. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which resulted in a variety of measurement uncertainty in the dataset, however, averaged over such a large number of stems this should not meaningfully alter the results of our analysis. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial, parameters estimated using a test dataset \citep{Ryanunbublished}.

@@

Plots were assigned to biogeographical groups, after \citet{Fayolle2018}. Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination. Plot data used in this study occurred in five biogeographical groups, see Table \ref{clust_tabl} for a description of each biogeographical cluster.

@@

Structural Equation Models (SEM) investigated the regional determinants of woody Above-Ground Biomass (AGB). All SEMs were constructed and analysed in the \textit{lavaan} package in R version 3.6.0. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}. A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is known to affect productivity, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, and soil composition. Indepedently of total precipitation, precipitation seasonality (measured as the covariance of daily precipitation) determines, simplistically, whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth.

We specified a conceptual model with latent variables based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity and tree structural diversity (Figure \ref{con_mod}). Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to soak up measurement error \citep{}.

@@

Observed variables of the latent variables were standardized to Z-scores for analysis. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates problems arising from the observed variables being on different scales. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix.

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention \citep{}. Due to some latent variables being regressed against both structural diversity and AGB, exact factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM and were therefore estimated by the SEM model.

For each biogeographical cluster, and at the regional level, we employed SEM to assess the relative importance of each latent variable in determining AGB. We assessed the indirect effects of abiotic environment on AGB via species and structural tree diversity to understand the interactive nature of species diversity and abiotic environment in determining AGB. In a model selection process, latent variable paths for each biogeographical cluster model were removed sequentially to investigate their removal on model fit. Model fit at each stage of the model selection process was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI) and the R\textsuperscript{2} coefficient of determination for AGB. The best models according to these goodness-of-fit statistics are presented in the results. See \todo{Appendix 1 for a full description of all models in the model description process.}

Two sets of SEMs were fitted. The first used only tree diversity latent constructs to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship. The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and AGB.

@@

Simple correlations showed a positive association between Aboveground Biomass (AGB) and a latent variable representing tree species diversity ($r_{(1522)}$ = 0.30, $t$ = 12.2). The tree structural diversity latent variable also had a positive association with AGB ($r_{(1327)}$ = 0.18, $t$ = 15.0).

In an SEM describing the indirect effect of tree species diversity on AGB via the mediating effect of stand structural diversity, species diversity had a negative effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}). Stand structural diversity had a strong positive effect on AGB (\todo{Test stats}). This effect was observed in biogeographical clusters 1, 2 and 3. Clusters 4 and 5 had small sample sizes due to missing data and so path regression uncertainties were high, with confidence intervals overlapping zero.

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes_clust_all}
	\caption{Path coefficients (slopes) for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Path coefficients are +/- 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be non-significant effects.}
	\label{struc_model_slopes}
\end{figure}

\input{include/struc_model_fit_clust_stats}

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{sem_struc_path}
	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
	\label{sem_struc_path}
\end{figure}

@@

The latent variable of tree species diversity had a weak positive effect on AGB across the region. A stronger positive effect was seen between structural diversity and aboveground biomass, with species diversity acting \textit{through} structural diversity to effect aboveground biomass.

@@

%\bibliography{/Users/johngodlee/google_drive/bib/lib}
